Liveness-Based Garbage Collection for Lazy Languages

### 2016-01-28 paper/paper.tex
more savings;

$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-1-clo} \\
\hline
$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, \heap,\,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, \heap,\, e_{\mathit{f}}$      &
{\sc funcall} \\
\hline
$\ell$ is a new location& {$\rho,\, S,\, \heap,\, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell := \langle s,
    \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
    \rangle],\, e$} &
{\sc let} \\
\hline
$\heap(\rho(x)) \ne 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S,\, \heap,\,  e_1$ & {\sc if-true} \\
\hline
$\heap(\rho(x)) = 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S, \,\heap, \, e_2$ & {\sc if-false} \\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, \heap, \, s$}
&
{\sc if-clo} \\
\hline
{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
e)\!:\!S,\, \heap,\,
(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$ &
{\sc return-whnf}\\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\, (\SRETURN~x)
  \rightsquigarrow$
$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, \heap,\,  s$} &
{\sc return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
%}\comment
\vspace*{-6pt}
\end{figure*}

@@

\input{fig-liveness}

  Closure  variables get  their  liveness values  through  a chain  of
  dependences beginning at  a variable at an evaluation  point.  As an
  example, consider the following program:

\input{example-to-illustrate-liveness}

The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
\leftarrow            (+~\py~1),\,           \py            \leftarrow
(\length~\px_{\mathit{\pi}_1})\rbrack$.  Indeed,  the chains of closures  in the
heap are a  runtime representation of these  dependences.  Since $\pz$
is evaluated  at $\psi_3$  due to  the expression  $\SRETURN~\pz$, the
demand made by the calling context(s)  of $f$ places a demand on $\pz$
which  will  impart a  liveness  to  $\px_{\pi_1}$.  Other  dependence
chains  which result  in  a liveness  for  $\px_{\pi_1}$ are  $\lbrack
\psi_1\!\!\!:\!\!\py   \leftarrow  (\length~\px_{\pi_1})\rbrack$   and
$\lbrack   \psi_2\!\!\!:\!\!\!\pw   \leftarrow  (/~\pu~\py),\,\,   \py
\leftarrow   (\length~\px_{\pi_1})\rbrack$.   The   liveness  analysis
described in this section declares the liveness of $\px_{\pi_1}$ to be
a union of  the liveness arising from these dependence  chains.  To be
safe, a garbage collection during  evaluation of $\py$ at $\psi_1$ has
to use  this liveness to  copy the heap starting from  $\px_{\pi_1}$.  However,
notice that if a garbage collection takes place while evaluating $\pz$
at $\psi_3$, it can safely consider only the liveness arising from the
dependence chain\linebreak  $\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow
(+~\py~1),\, \py \leftarrow (\length~\px_{\pi_1})\rbrack$.  The garbage
collection  scheme  described  in Section~\ref{sec:GC-scheme}  uses  a
generalization of this observation  to dynamically select a evaluation
point specific liveness in order to collect more garbage.

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.   Here are  the ideas  behind the
proof.
\begin{enumerate}
\item       We   augment     the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a dynamic liveness analysis 
that records liveness values at each reference created during execution.
Any attempt to dereference a location with a null-liveness  
  during  a  transition will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline1} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{enumerate}

@@

To set up the minefield semantics, we follow these steps:
\begin{enumerate}

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environments for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

@@

In  summary, the  equations  generated during  liveness
analysis are:
\begin{enumerate}
\item   For  each   function  $\mathit{f}$,   equations
  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item   For  each   function  $\mathit{f}$   (including
  $\mainpgm$) an  equation  defining
  liveness for each garbage in  $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}      The
equations  above   can  now  be  re-interpreted   as  a
context-free    grammar   (CFG)    on   the    alphabet
$\lbrace\acar,  \acdr,  \bcar,  \bcdr,  \clazy\rbrace$.
Let \var{$X$}  denote the  non-terminal for  a variable
$X$  occurring on  the LHS  of the  equations generated
from  the  analysis.  We  can  think  of the  resulting
productions as being  associated with several grammars,
one for each non-terminal \var{\Lanv{x}{}} regarded as
a  start  symbol.   As  an  example,  the  grammar  for
\var{\Lanv{\xl}{}} and \var{\Lanv{\pa}{}}    comprise     the     following
productions:
\begin{eqnarray*}
  \var{\Lanv{\xl}{}}  &\rightarrow& 
  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
\var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{x}{}},   denoted
$\mathscr{L}(\var{\Lanv{x}{}})$,   is    the   desired
solution  of  \Lanv{x}{}.    
However, recall that the decision problem that we are interested in
during garbage collection is the following: 
\begin{quote}
Let $x.\alpha$ be a forward access path (consisting only
of edges  \acar\ and \acdr\  but not \bcar,  \bcdr\ or
\clazy).       Let      $\mathscr{L}(\var{\Lanv{x}{}})
\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
consists  of  forward  access  paths  only.  Then  does
$\alpha \in \sigma$?
\end{quote}

We now model  the above problem as one  of deciding the
membership of  a context-free grammar augmented  with a
fixed set of unrestricted productions.

@@

\begin{enumerate}
\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
  are two productions
  $\nont{i}{c} \rightarrow \overline{c'} \nont{j}{\acar}\bcar$ and $\nont{i}{c}
  \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
\end{enumerate}
The idea  behind the  productions is explained  with an
example:  Assume that  the current  sentential form  is
$\bcar\bcdr\nont{i}{\acar}\acar\acar\$$. Also  assume that  the
TM  has  a   transition  $(S_i,\acar)  \rightarrow
(S_j,\acdr, L)$.  Since the next corresponding  step in
the  derivation  has  to  be  done  without  any  prior
knowledge of whether the symbol to the left of the tape
is a \acar\  or a \acdr, two  productions are provided,
and  the  invariant  will  be maintained  only  if  the
production         $\nont{i}{\acar}         \rightarrow
\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
in  the  derivation.    This  gives  the  configuration
$\bcar\bcdr\acdr\nont{j}{\acdr}\acdr\acar\acar\$$.
Simplification   with    the   production   $\bcdr\acdr
\rightarrow               \epsilon$              yields
$\bcar\nont{j}{\acdr}\acdr\acar\acar\$$,    which   exactly
corresponds  to   the  changed  configuration   of  the
TM.  Notice  carefully that a wrong  choice breaks
the invariant  and it cannot be  recovered subsequently
by any choice of productions.

After the  TM has  halted, there  are further  ``cleanup'' derivations
that derive  $\epsilon$ only if  the invariant has been  maintained so
far.    For   every   symbol   $c$,  we   introduce   a   non-terminal
$\nont{\mathit{final}}{c}$  where   $\nont{\mathit{final}}{}$  is  the
final state  of the  TM. We add  productions $\nont{\mathit{final}}{c}
\rightarrow         \acar        \nont{\mathit{final}}{c}$         and
$\nont{\mathit{final}}{c} \rightarrow  \acdr \nont{\mathit{final}}{c}$
for cleaning up the \bcar\ and \bcdr\  symbols on the left of the head
and                $\nont{\mathit{final}}{c}               \rightarrow
\nont{\mathit{final}}{c}\bcar$      and      $\nont{\mathit{final}}{c}
\rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
\acdr\ on the right of the tape head.  This completes the
reduction.\\ \mbox{}
\end{proof}

@@

We circumvent  the problem  of undecidability  by over-
approximating the CFG by non-deterministic finite state
automata (NFA). The NFAs  are then simplified using the
$\hookrightarrow$ rules.   Finally the  simplified NFAs
are converted to DFAs.

@@

\section{The garbage collection scheme}
\label{sec:GC-scheme}

@@

Based on  the above considerations,  we restrict the  possible garbage
collection points in a function body to the following:
\begin{enumerate}
\item We  maintain static estimates  of the memory required  to create
  the closures for each function body.  On entering a function, if the
  available memory is  less than this requirement for  the function, a
  garbage collection is triggered.
\item Since the evaluation of a  \SIF\ condition may trigger a 
  collection, after evaluating the condition   the available memory
  is  checked once  again against  a  revised estimate  of the  memory
  (based on  value of the condition)  required to execute the  rest of
  the program.  A garbage collection is triggered if enough memory is unavailable.
\end{enumerate}

@@

\section{Related work} 
\label{sec:relatedwork}
The impact of  liveness on the effectiveness of  garbage collection is
investigated in~\cite{Hirzel}.   The observation is that  liveness can
significantly  impact   garbage  collection,  but  only   when  it  is
interprocedural.  As far as memory requirement is concerned, our paper
demonstrates this observation.

@@

Simplifiers~\cite{ONeill} are abstractly described as lightweight
daemons that attach themselves to program data and, when activated, improve
the efficiency of the program. Our liveness based garbage collection
as an instance of a  simplifier which is tightly coupled with garbage collectors.  The
approach that is closest to the  method described in this paper is the
liveness-based garbage collector implemented
in~\cite{karkare07liveness,asati14lgc} and address eager languages. We
extend this to handle lazy evaluation and closures.

>>>>

$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-1-clo} \\
\hline
$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, \heap,\,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, \heap,\, e_{\mathit{f}}$      &
{\sc funcall} \\
\hline
$\ell$ is a new location& {$\rho,\, S,\, \heap,\, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell := \langle s,
    \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
    \rangle],\, e$} &
{\sc let} \\
\hline
$\heap(\rho(x)) \ne 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S,\, \heap,\,  e_1$ & {\sc if-true} \\
\hline
$\heap(\rho(x)) = 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S, \,\heap, \, e_2$ & {\sc if-false} \\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, \heap, \, s$}
&
{\sc if-clo} \\
\hline
{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
e)\!:\!S,\, \heap,\,
(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$ &
{\sc return-whnf}\\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\, (\SRETURN~x)
  \rightsquigarrow$
$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, \heap,\,  s$} &
{\sc return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
%}\comment
\vspace*{-16pt}
\end{figure*}

@@

\begin{figure}[t]
  \input{example-to-illustrate-liveness}  
  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
\end{figure}
\input{fig-liveness}

Closure  variables get  their  liveness values  through  a chain  of
dependences beginning at  a variable at an evaluation  point.  As an
example, consider the program in Figure~\ref{fig:lv-closure}.
The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
\leftarrow            (+~\py~1),\,           \py            \leftarrow
(\length~\px_{\mathit{\pi}_1})\rbrack$.  Indeed,  the chains of closures  in the
heap are a  runtime representation of these  dependences.  Since $\pz$
is evaluated  at $\psi_3$  due to  the expression  $\SRETURN~\pz$, the
demand made by the calling context(s)  of $f$ places a demand on $\pz$
which  will  impart a  liveness  to  $\px_{\pi_1}$.  Other  dependence
chains  which result  in  a liveness  for  $\px_{\pi_1}$ are  $\lbrack
\psi_1\!\!\!:\!\!\py   \leftarrow  (\length~\px_{\pi_1})\rbrack$   and
$\lbrack   \psi_2\!\!\!:\!\!\!\pw   \leftarrow  (/~\pu~\py),\,\,   \py
\leftarrow   (\length~\px_{\pi_1})\rbrack$.   The   liveness  analysis
described in this section declares the liveness of $\px_{\pi_1}$ to be
a union of  the liveness arising from these dependence  chains.  To be
safe, a garbage collection during  evaluation of $\py$ at $\psi_1$ has
to use  this liveness to  copy the heap starting from  $\px_{\pi_1}$.  However,
notice that if a garbage collection takes place while evaluating $\pz$
at $\psi_3$, it can safely consider only the liveness arising from the
dependence chain\linebreak  $\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow
(+~\py~1),\, \py \leftarrow (\length~\px_{\pi_1})\rbrack$.  The garbage
collection  scheme  described  in Section~\ref{sec:GC-scheme}  uses  a
generalization of this observation  to dynamically select a evaluation
point specific liveness in order to collect more garbage.

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.   Here are  the ideas  behind the
proof.
\begin{myenumerate}
\item       We   augment     the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a dynamic liveness analysis 
that records liveness values at each reference created during execution.
Any attempt to dereference a location with a null-liveness  
  during  a  transition will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline1} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{myenumerate}

@@

To set up the minefield semantics, we follow these steps:
\begin{myenumerate}

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{myenumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environments for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{myenumerate}
\end{myenumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

@@

In  summary, the  equations  generated during the
analysis are:
\begin{myenumerate}
\item   For  each   function  $\mathit{f}$,   equations
  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item   For  each   function  $\mathit{f}$   (including
  $\mainpgm$) an  equation  defining
  liveness for each garbage in  $e_{\mathit f}$.
\end{myenumerate}

\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}

The  equations  above can  now  be  re-interpreted as  a  context-free
grammar  (CFG) on  the  alphabet $\lbrace\acar,  \acdr, \bcar,  \bcdr,
\clazy\rbrace$.  Let \var{$X$} denote  the non-terminal for a variable
$X$ occurring on the LHS of the equations generated from the analysis.
We can  think of  the resulting productions  as being  associated with
several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
as a start symbol.  As  an example, the grammar for \var{\Lanv{\xl}{}}
and \var{\Lanv{\pa}{}} comprise the following productions:
\begin{eqnarray*}
  \var{\Lanv{\xl}{}}  &\rightarrow& 
  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
\var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{x}{}},   denoted
$\mathscr{L}(\var{\Lanv{x}{}})$,   is    the   desired
solution  of  \Lanv{x}{}.    
However, recall that the decision problem that we are interested in
during garbage collection is: 
%%\begin{quote} 
{\em
Let $x.\alpha$ be a forward access path (consisting only
of edges  \acar\ and \acdr\  but not \bcar,  \bcdr\ or
\clazy).       Let      $\mathscr{L}(\var{\Lanv{x}{}})
\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
consists  of  forward  access  paths  only.  Then  does
$\alpha \in \sigma$?
}
%%\end{quote}
We model  this problem as one  of deciding the
membership of  a CFG augmented  with a
fixed set of unrestricted productions.

@@

\begin{myenumerate}
\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
  are two productions
  $\nont{i}{c} \rightarrow \overline{c'} \nont{j}{\acar}\bcar$ and $\nont{i}{c}
  \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
\end{myenumerate}
The idea  behind the  productions is explained  with an
example:  Assume that  the current  sentential form  is
$\bcar\bcdr\nont{i}{\acar}\acar\acar\$$. Also  assume that  the
TM  has  a   transition  $(S_i,\acar)  \rightarrow
(S_j,\acdr, L)$.  Since the next corresponding  step in
the  derivation  has  to  be  done  without  any  prior
knowledge of whether the symbol to the left of the tape
is a \acar\  or a \acdr, two  productions are provided,
and  the  invariant  will  be maintained  only  if  the
production         $\nont{i}{\acar}         \rightarrow
\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
in  the  derivation.    This  gives  the  configuration
$\bcar\bcdr\acdr\nont{j}{\acdr}\acdr\acar\acar\$$.
Simplification   with    the   production   $\bcdr\acdr
\rightarrow               \epsilon$              yields
$\bcar\nont{j}{\acdr}\acdr\acar\acar\$$,    which   exactly
corresponds  to   the  changed  configuration   of  the
TM.  Notice  carefully that a wrong  choice breaks
the invariant  and it cannot be  recovered subsequently
by any choice of productions.

After the  TM has  halted, there  are further  ``cleanup'' derivations
that derive  $\epsilon$ only if  the invariant has been  maintained so
far.    For   every   symbol   $c$,  we   introduce   a   non-terminal
$\nont{\mathit{final}}{c}$  where   $\nont{\mathit{final}}{}$  is  the
final state  of the  TM. We add  productions $\nont{\mathit{final}}{c}
\rightarrow         \acar        \nont{\mathit{final}}{c}$         and
$\nont{\mathit{final}}{c} \rightarrow  \acdr \nont{\mathit{final}}{c}$
for cleaning up the \bcar\ and \bcdr\  symbols on the left of the head
and                $\nont{\mathit{final}}{c}               \rightarrow
\nont{\mathit{final}}{c}\bcar$      and      $\nont{\mathit{final}}{c}
\rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
\acdr\ on the right of the tape head.  This completes the
reduction.\!
\end{proof}

@@

@@

\vspace*{-2.5mm}
\section{The garbage collection scheme}
\label{sec:GC-scheme}

@@

Based on  the above considerations,  we restrict the  possible garbage
collection points in a function body to the following:
\begin{myenumerate}
\item We  maintain static estimates  of the memory required  to create
  the closures for each function body.  On entering a function, if the
  available memory is  less than this requirement for  the function, a
  garbage collection is triggered.
\item Since the evaluation of a  \SIF\ condition may trigger a 
  collection, after evaluating the condition   the available memory
  is  checked once  again against  a  revised estimate  of the  memory
  (based on  value of the condition)  required to execute the  rest of
  the program.  A garbage collection is triggered if enough memory is unavailable.
\end{myenumerate}

@@

\vspace*{-2mm}
\section{Related work} 
\label{sec:relatedwork}
The impact of  liveness on the effectiveness of  garbage collection is
investigated in~\cite{Hirzel}.   The observation is that  liveness can
significantly  impact   garbage  collection,  but  only   when  it  is
interprocedural.  As far as memory requirement is concerned, our paper
demonstrates this observation.

@@

Simplifiers~\cite{ONeill} are abstractly described as lightweight
daemons that attach themselves to program data and, when activated, improve
the efficiency of the program. Our liveness-based garbage collection
as an instance of a  simplifier which is tightly coupled with garbage collectors.  The
approach that is closest to the  method described in this paper is the
liveness-based garbage collector implemented
in~\cite{karkare07liveness,asati14lgc} and address eager languages. We
extend this to handle lazy evaluation and closures.
